Multi-domain, multilingual translation with latent multi-task group dropout

### 2021-11-09 2022/acl/acl_latex-fy.tex
intro rewrite

\newcommand{\fyTodo}[1]{\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyTodostar}[1]{\Todo*[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyDone}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyFuture}[1]{\done[FY]\Todo[FY:]{\textcolor{red}{#1}}}
\newcommand{\fyDonestar}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\jcTodo}[1]{\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\jcDone}[1]{\done[JC]\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\revision}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondone}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondel}[1]{}
\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{{#1}}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}

@@

\begin{document}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
\maketitle
\begin{abstract}
Multi-domain and multilingual translation can be cast under multi-task learning in which each domain or language pair is considered a task. Sharing most of the parameters of the model causes a catastrophic interference between unrelated tasks. Most of the proposed multi-task architectures propose sharing the parameters of an underlying generic model and using different output layers to predict different tasks. However, these paradigms do not explicitly exploit the proximity between tasks. In this work, we aim to express the task proximity in the multi-task network. Instead of using a predefined multi-task network, we jointly learn routing task-dependent sub-networks and the parameters of the whole network. More precisely, we propose clustering the nodes of each layer into equal-sized groups. For each layer and each domain, a number of groups will be dropped. By consequence, we assign different sparse sub-networks to different tasks. Our method learns the task-dependent dropping selections and the parameters of the underlying network. Our approach allows the proximity between tasks to express in their dropping selections. Through an extensive set of experiments, we demonstrate the advantages of our method over strong baselines in multi-domain and multilingual translation. Furthermore, through an ablation study, we illustrate a significant correlation between the similarity between learned task-dependent dropping selections and the proximity between tasks.
\end{abstract}

\begin{abstract}
  [FY] Multi-domain and multi-lingual machine translation often rely on parameter sharing strategies, where large portions of the network are meant to capture the commonalties of the task at hand, while smaller parts are reserved to model the peculariaties of a language or a domain. In adapter-based approaches, these strategies are hard coded in the network architecture. In this work, we propose a new method aimed at learning the parameter sharing pattern, using a latent-variable model. We show how this model can be trained end to end, and report experimental results which show that the learned pattern are both meaningful, and yield improved translation performance without any increase of the model size. 
\end{abstract}

\section{Introduction}
Multi-domain and multilingual machine translation (MMT) aim to develop one single model to perform translations for multiple domains and multiple language pairs, respectively. These paradigms are motivated by the compactness of the resulting translation system \citep{dabre20survey,Chu18multilingual}, the hypothetical positive knowledge transfer between similar domains \citep{Pham21revisiting} or between languages in the same family \citep{Tan19multilingual}. However, having all the tasks use exactly the same model parameters can cause negative interference between unrelated tasks \citep{conneau20unsupervised,wang20negative}. Hence the recent development of approaches relying on a partial sharing of the parameters, eg.\ using the adapter layer of \citet{Bapna19simple,houlsby19parameter,Pham20Study}.

In multi-domain setting, \citet{Pham21revisiting} pointed out a challenge in how multi-domain systems fully exploit the similarity between domains. The authors demonstrated the weakness of these systems in handling the proximity between domains through a simple setting using a pseudo domain separation. Domain adapters, which are strong baselines for multi-domain translation \cite{Pham20Study}, failed this test. Such predefined allocations of parameters to domains do not take into account the proximity of domains in their choice.

The method we developed here mitigates the catastrophic interference in training a neural machine translation (NMT) model for multiple domains or multiple language pairs. Our main contribution are summarized below:
\begin{itemize}
\setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
	\item We provide a rigorous mathematical modeling the problem of jointly learning routing task-dependent sub-network and the parameters of the underlying model using variational probabilistic modeling.
	\item Our method propose end to end training with very little extra computational cost.
	\item With an extensive experimental setting, we demonstrate a significant gain of our method against strong baselines.
	\item We illustrate a significant correlation between the similarity between learned sub-networks and the proximity of their corresponding tasks.
\end{itemize}

>>>>

\newcommand{\fyTodo}[1]{\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyTodostar}[1]{\Todo*[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyDone}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyFuture}[1]{\done[FY]\Todo[FY:]{\textcolor{red}{#1}}}
\newcommand{\fyDonestar}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\jcTodo}[1]{\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\jcDone}[1]{\done[JC]\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\revision}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondone}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondel}[1]{}
\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{{#1}}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}
\newcommand{\mtsquare}[0]{MT$^2$}

@@

\begin{document}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
\maketitle
\begin{abstract}
Multidomain and multilingual translation can be cast under multi-task learning in which each domain or language pair is considered a task. Sharing most of the parameters of the model causes a catastrophic interference between unrelated tasks. Most of the proposed multi-task architectures propose sharing the parameters of an underlying generic model and using different output layers to predict different tasks. However, these paradigms do not explicitly exploit the proximity between tasks. In this work, we aim to express the task proximity in the multi-task network. Instead of using a predefined multi-task network, we jointly learn routing task-dependent sub-networks and the parameters of the whole network. More precisely, we propose clustering the nodes of each layer into equal-sized groups. For each layer and each domain, a number of groups will be dropped. By consequence, we assign different sparse sub-networks to different tasks. Our method learns the task-dependent dropping selections and the parameters of the underlying network. Our approach allows the proximity between tasks to express in their dropping selections. Through an extensive set of experiments, we demonstrate the advantages of our method over strong baselines in multi-domain and multilingual translation. Furthermore, through an ablation study, we illustrate a significant correlation between the similarity between learned task-dependent dropping selections and the proximity between tasks.
\end{abstract}

\begin{abstract}
  [FY] Multidomain and multilingual machine translation often rely on parameter sharing strategies, where large portions of the network are meant to capture the commonalties of the task at hand, while smaller parts are reserved to model the peculariaties of a language or a domain. In adapter-based approaches, these strategies are hard coded in the network architecture. In this work, we propose a new method aimed at learning the parameter sharing pattern, using a latent-variable model. We show how this model can be trained end to end, and report experimental results which show that the learned pattern are both meaningful, and yield improved translation performance without any increase of the model size. 
\end{abstract}

\section{Introduction}
Multidomain and multilingual machine translation aim to develop one single model to perform translations for multiple domains and multiple language pairs, respectively.\footnote{We will refer to these two situations as 'multi-task MT' (or \mtsquare{} for short), and refer to individual domains and languages as 'tasks'.} These paradigms are motivated by the compactness of the resulting translation system \citep{dabre20survey,Chu18multilingual}, the hypothetical positive knowledge transfer between similar domains \citep{Pham21revisiting} or between languages in the same family \citep{Tan19multilingual}. However, having all the tasks use exactly the same model parameters can cause negative interference between unrelated tasks \citep{conneau20unsupervised,wang20negative}. Hence the recent development of approaches relying on a partial sharing of the parameters, eg.\ using adapter layers as studied in \citet{houlsby19parameter,Bapna19simple,Pham20Study}. If these techniques have proven effective to achieve strong baselines, such fail to fully take advantage of the similarities that exist between domains and tasks, as documented eg.\ in \citep{Pham21revisiting}. This is because the partition of the parameter space between generic or task-specific sub-parts, and their allocation to each task is hard coded in the network, irrespective of the existing commonalties and differences in the data space.

In this paper, we try to develop techniques that can take the similarity between tasks into account in a more effective way, by learning the network organization from the data. To this end, we introduce a set of latent variables in the model, to account for the unseen association between tasks and regions of the parameter space, and show how training can still be performed end-to-end using a variational surrogate of the standard loss function. Our experiments with multilingual and multidomain confirm that this method can automatically detect similarities in the data, and yield related tasks to share more parameters. They also show that this method is comparable to using adapter layers in a number of empirical comparisons; contrarily to adapters, these performance are obtained without any increase of the model size.  Our main contribution are summarized below:
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
\item We provide a rigorous mathematical formulattion to the problem of jointly learning routing task-dependent sub-networks and the parameters of the underlying models using variational probabilistic modeling;
\item We introduce algorithms to train this model end to end with very little extra computational cost.
\item We report, using an extensive experimental setting, significant gains of this method with respect to strong baselines.
\item We study how this method can actually exploit the similarities between tasks to learn consistent sub-networks.
\end{itemize}

\fyTodo{Can you do distillation ? That is only keep the parameter for one domain and still get good results (better than DA ?)}
